---
title: ''
output:
  powerpoint_presentation:
    reference_doc: ub_template.pptx
  slidy_presentation: default
  ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# ---- Library Imports ----
library(RSQLite)
library(DBI)
library(ggplot2)
library(kableExtra)
library(tidyverse)
library(cluster)
library(magrittr)
library(plotly)
library(caret)
library(ggbiplot)
library(cowplot)
library(janitor)
library(patchwork)
library(ggdendro)
library(ISLR2)
library(datawizard)
library(ggplot2)
library(dendextend)
library(data.table)
library(factoextra)
library(graphics)

# ---- Setting seed = 1 for reproducibility ----
set.seed(1)
```

```{r Data pull, echo = FALSE, message = FALSE, warning = FALSE}
# SQL Data base connection & data pull
conn <- dbConnect(RSQLite::SQLite(), "SpotifyData.db")      # This line connects to our database

# Pulling data into environment
sql <- "
-- This data was used to pull information on tracks and what countries they are popular in for a project on Clustering!
WITH count_track_nbr_of_countries AS(
    SELECT DISTINCT
  	  track_id
      ,COUNT(name) AS num_countries
    FROM Playlists
    GROUP BY track_id
  )
  
SELECT  
	Playlists.name
	,count_track_nbr_of_countries.num_countries
	,Tracks.duration_ms
	,Tracks.explicit
	,Tracks.current_popularity
	,Tracks.acousticness
	,Tracks.danceability
	,Tracks.instrumentalness
	,Tracks.key_signature
	,Tracks.mode
	,Tracks.tempo
	,Tracks.time_signature
	,Tracks.valence
FROM Playlists
JOIN Tracks
	ON Tracks.track_id = Playlists.track_id
JOIN count_track_nbr_of_countries
	ON count_track_nbr_of_countries.track_id = Playlists.track_id;
"

# Get data
data <- dbGetQuery(conn, sql)
data$mode <- as.integer(data$mode)
```

```{r Create Standardized Data, echo = FALSE, message = FALSE, warning = FALSE}
# Standardized
data_std <- standardize(data)
```

# Group 1 - Clustering Methods for Spotify Song Data
Jason Caballes

Nick Cruz

Sean Grzenda

Karthik Sri Sakthi Degala

# Objective 
Our goal is to perform clustering algorithms, i.e. K-means and hierarchical K-means to 
try and  unveil potential similarities among the most popular songs today across the globe. 
We want to see if favorable music from each country will naturally form clusters.

We may cluster tracks based on their musical attributes to see if tracks within the clusters 
are more popular in certain countries. Clustering tracks may also allow us to generate playlists,
such as Spotify’s own recommendation systems


# Methods Used 
Created a database file containing the top tracks from 67 countries

we explored the following for analysis: 

* Hierarchical Clustering  and Dendrograms  
* Elbow method
* Gap Statistic
* Silhouette
* K-Means Clustering
* PCA to provide a two-dimensional visualization of the clusters


# Hierarchical Clustering and Dendrograms

```{r, echo = FALSE, message = FALSE, warning = FALSE}

# Find distances
x_dist <- dist(data)

# Complete method
hc_complete <- hclust(x_dist, method = "complete")

# Average Method
hc_average <- hclust(x_dist, method = "average")

# Initial Dendrogram
linkage_dendro <- ggdendrogram(hc_complete, rotate=T, labels=F) + ggtitle('Complete Linkage Dendrogram')
average_dendro <- ggdendrogram(hc_average, rotate=T, labels=F) + ggtitle('Average Linkage Dendrogram')

# Create subplot
dendros <- linkage_dendro / average_dendro 

# Output dendrograms
dendros
```


# Elbow Method to find Optimal Clusters


Elbow Method Overview

1. Identifies optimal clusters in k-means by locating the "elbow" point.
2. Elbow point: Where WCSS decrease shifts markedly.

Analysis of Spotify Data

1. No clear elbow observed in the plot.
2. Implication: Data may not naturally form distinct clusters.
3. Possible complexity in the clustering structure.


# Elbow Method to find Optimal Clusters

```{r Elbow Method, echo = FALSE, message = FALSE, warning = FALSE}
# Remove name field as we only require numerical data
data_no_strings <- data |> select(-c("name"))

# Scale the data
df_scale <- scale(data_no_strings)

# Create Elbow Plots
km_out_list <- lapply(1:20, function(k) list(
  k=k,
  km_out=kmeans(df_scale, k, nstart = 20)))

km_results <- data.frame(
  k=sapply(km_out_list, function(k) k$k),
  totss=sapply(km_out_list, function(k) k$km_out$totss),
  tot_withinss=sapply(km_out_list, function(k) k$km_out$tot.withinss)
  )

fig <- plot_ly(km_results,x=~k,y=~tot_withinss) |> add_markers() |> add_paths() |> layout(title="Elbow Plot")
fig
```


# Gap statistic to Find optimal Clusters

Gap Statistic Overview

1. Measures optimal clusters by comparing within-cluster variation to a null reference.
2. Assesses against expected values under null distribution.

Optimal Clusters Identified

1. Gap statistic suggests optimal clusters: 2, 4, 9, 16.
2. Explore each cluster solution's impact on data insights.
3. Assess how different cluster counts influence predictions.



# Gap Statistic to find Optimal Clusters
```{r Gap, echo = FALSE, message = FALSE, warning = FALSE}
gap_kmeans <- clusGap(df_scale, kmeans, nstart = 20, K.max = 20, B = 20)

plot(gap_kmeans, main = "Gap Statistic: kmeans")
```


# Silhouette to find Optimal Clusters

Silhouette Method Overview

1. Calculates the silhouette score for each data point.
2. Measures closeness within a cluster versus neighboring clusters.

Optimal Cluster Numbers

1. Silhouette method identifies optimal clusters: 8, 13, 19.
2. Higher silhouette scores at these clusters.
3. Indicates better separation and cohesion among clusters.


# Silhouette to find Optimal Clusters
```{r Silhouette, echo = FALSE, message = FALSE, warning = FALSE}
fviz_nbclust(df_scale, kmeans, method = "silhouette", nboot = 20, k.max=20, nstart=20, iter.max=40) + labs(subtitle = "Silhouette method")
```


# PCA

PCA simplifies data by transforming related variables into uncorrelated ones.
Resulting components capture key patterns and variance.
Ideal for reducing complexity in large datasets.


# PCA

```{r PCA, echo = FALSE, message = FALSE, warning = FALSE}
# Run PCA on df_scale, make biplot and scree plot/percentage variance explained plot
# save as pc_out, we will use pc_out$x[,1] and pc_out$x[,2] later for plotting

pc.out <- prcomp(subset(data_std, select = -name))

# biplot
#biplot(pc.out,scale=0)

# skree plot
PVE <- tibble(
  PC=1:length(pc.out$sdev),
  Var=pc.out$sdev^2,
  PVE=Var/sum(Var),
  CumPVE=cumsum(PVE)
)
cowplot::plot_grid(
qplot(data=PVE,x=PC,y=PVE,geom=c("point","line"),
      xlab = "Principal Component",
      ylab = "Proportion of Variance Explained"),
qplot(data=PVE,x=PC,y=CumPVE,geom=c("point","line"),
      xlab = "Principal Component",
      ylab = "Cumulative Proportion of Variance Explained")
) 
```


# PCA Plot of K-Means Clusters

```{r Plotting Clusters, echo = FALSE, message = FALSE, warning = FALSE}
# ---- Running the k-means algorithms based of optimal K from metrics ----

# Applying k means with 67 clusters
# This is is an attempt to mimic the 67 playlists, but with k-means
km.out.67 = kmeans(df_scale,67)
df_clustered_67=as.data.frame(df_scale)
df_clustered_67$cluster = as.factor(km.out.67$cluster)

# Silhouette
km.out.19 = kmeans(df_scale,19)
df_clustered_19=as.data.frame(df_scale)
df_clustered_19$cluster=as.factor(km.out.19$cluster)

# Gap statistics
km.out.9 = kmeans(df_scale,9)
df_clustered_9=as.data.frame(df_scale)
df_clustered_9$cluster=as.factor(km.out.9$cluster)

# elbow 
km.out.5 = kmeans(df_scale,5)
df_clustered_5=as.data.frame(df_scale)
df_clustered_5$cluster = as.factor(km.out.5$cluster)

# ---- Creating figures ----
playlist_plot <- ggbiplot(pc.out, scale=0, groups=data$name, var.axes=F,
                          point.size=0.5) |> ggplotly() |> layout(showlegend = FALSE)

k_means_67_plot <- ggbiplot(pc.out, scale=0, groups=df_clustered_67$cluster, 
                  var.axes=F, point.size=0.5) |> ggplotly() |> layout(showlegend = FALSE)

silhouette_plot <- ggbiplot(pc.out,scale=0, groups=df_clustered_19$cluster,
                  var.axes=F, point.size=0.5) |> ggplotly() |> layout(showlegend = FALSE)

gap_plot <- ggbiplot(pc.out,scale=0, groups=df_clustered_9$cluster,
                var.axes=F, point.size=0.5) |> ggplotly() |> layout(showlegend = FALSE)

elbow_plot <- ggbiplot(pc.out, scale=0, groups=df_clustered_5$cluster,
                  var.axes=F, point.size=0.5) |> ggplotly() |> layout(showlegend = FALSE)

```


```{r Plotting k-means vs original data, echo = FALSE, message = FALSE, warning = FALSE}
subplot(playlist_plot, k_means_67_plot, nrows=2, shareX=T) |>
  layout(annotations=list(
    list(y=1.0, text="Plot of Tracks and their Countries", showarrow=F, xref='paper', yref='paper'), 
    list(text="Plot of K-Means to Group Tracks", showarrow=F, xref='paper', yref='paper'))
  ) 
```

\

```{r various optimal Ks, echo = FALSE, message = FALSE, warning = FALSE}
subplot(silhouette_plot, elbow_plot, gap_plot, nrows=2, shareX=T) |>
  layout(annotations=list(
    list(x=0.2, y=1.0, text="Silhouette", showarrow=F, xref='paper', yref='paper'), 
    list(x=0.8, y=1.0, text="Elbow", showarrow=F, xref='paper', yref='paper'),
    list(x=0.2, text="Gap", showarrow=F, xref='paper', yref='paper'))
    )
```


# Distribution of countries Between clusters

```{r barPlots, echo = FALSE, message = FALSE, warning = FALSE}
df_clustered_5$countries <- data$name

figure <- ggplot(df_clustered_5, aes(x = as.factor(cluster), fill = countries)) + geom_bar() + theme(legend.position = 'none') + xlab("Cluster") + ylab("Number of instances of a country")
ggplotly(figure)
```


# Summary


1. Gap statistics identified optimal clusters as 2, 4, 9, and 16.
2. Silhouette method suggested optimal clusters as 8, 13, and 19.
3. Chose to test clusters at 5, 9, 19 and 67 (total countries).
4. Clusters at 5 and 9 showed decent differentiation on PCA plots.
5. Cluster at 19 provided good results with visible grouping, though not all clusters were distinct.
6. Cluster at 67 (equal to the number of countries) performed poorly, mirroring the initial country-based results.

# Conclusion

The answer to the original question is unfortunately no, songs do not cluster by country. In fact, the optimal number of clusters is nowhere close to the number of countries. 